Icicle: write once, run once

### 2016-05-30 2016fhpc/section/02-Source.tex
Merge conflicts

The grammar for Icicle is given in figure~\ref{fig:source:grammar}.
Value types, $T$, can be numbers, booleans or maps.
Modal types, $\TauMode$ can be a pure value type, or a modality associated with a value type.
Function types, $\TauFun$, are a list of argument types, and a single return type.
As Icicle is a first-order language, function types are not value types.

$\mi{Table}$ defines a table with a name and the names and types of columns.
$\mi{Exp}$ defines the expressions used as query and function bodies.
Primitive application is defined as prefix, but we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.
$\mi{Prim}$ defines the primitive types.
Function and query definitions are given in $\mi{Def}$.
$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.
All queries in a $\mi{Top}$ are fused into a single loop over the input data.

We now take a more detailed look at the group example from \S2, as these versions did not define the functions @last@ and @mean@.
We write a function @count@, which takes no arguments.
Its definition is a fold that starts at zero, and increments by one for every element.
\begin{code}
  function count
   = fold c = 0 then c + 1;
\end{code}

Next, @sum@ takes one argument, which is the elements to compute the sum of.
Its definition is a fold that starts at zero, and increments by the sum element for each element.
We can then define @mean@ as the sum divided by the count.
\begin{code}
  function sum  (v : Element Real)
   = fold s = 0 then s + v;
  function mean (v : Element Real)
   = sum v / count;
\end{code}

Finally, the @last@ function takes the elements of keys, and for every element returns that element.
Here we use an arbitrary initial value of 1900-01-01, but in reality this would return an optional value.
\begin{code}
  function last (k : Element Date)
   = fold l = date 1900 1 1 then k;
\end{code}

% Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.
% The query operations can define a fold such as a sum or a count, filtering according to a predicate, or grouping by a key.
% Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
% Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.

@@

% ---------------------------------------------------------
\subsection{Type system}
The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.
Judgment $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ means that under context $\Gamma$, the expression has a given value type and modality.
Judgment $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
Judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
Judgment $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.

>>>>

The grammar for Icicle is given in Figure~\ref{fig:source:grammar}. Value types $T$ include numbers, booleans or maps. Modal types $\TauMode$ include the pure value types, and modalities associated with a value type. Function types $\TauFun$ include non-function modal types, or a function from modal type to function type. As Icicle is a first-order language, function types are not value types.

Table definitions $\mi{Table}$ define a table name and the names and types of columns. Expressions $\mi{Exp}$ can be variable names, constants, applications of primitives or functions. The @fold@ construct defines the name of an accumulator, the expression for the initial value, and the expression used to update the accumulator for each element of the stream. The @filter@ construct defines a predicate and an expression to accumulate values for which the predicate is true. The @group@ construct defines an expression used to determine the key for each element of the stream, and an expression to accumulate the values that share a common key.

$\mi{Prim}$ defines the primitive operators.
Function and query definitions are given in $\mi{Def}$. $\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries. All queries in a top-level program process the same table.

% We now take a more detailed look at the group example from \S2, as these versions did not define the functions @last@ and @mean@. 
% We write a function @count@, which takes no arguments.
% Its definition is a fold that starts at zero, and increments by one for every element.
% \begin{code}
%   function count
%    = fold c = 0 then c + 1;
% \end{code}

% Next, @sum@ takes one argument, which is the elements to compute the sum of.
% Its definition is a fold that starts at zero, and increments by the sum element for each element.
% We can then define @mean@ as the sum divided by the count.
% \begin{code}
%   function sum  (v : Element Real)
%    = fold s = 0 then s + v;
%   function mean (v : Element Real)
%    = sum v / count;
% \end{code}

% Finally, the @last@ function takes the elements of keys, and for every element returns that element.
% Here we use an arbitrary initial value of 1900-01-01, but in reality this would return an optional value.
% \begin{code}
%   function last (k : Element Date)
%    = fold l = date 1900 1 1 then k;
% \end{code}

% Expressions can be variable names, primitive operators such as addition and division, function application, let bindings, or query operations.

% a predicate, or grouping by a key.
% Function application $\mi{x~Exp*}$ is the name of a top-level function applied to any number of arguments.
% Primitive application $\mi{Prim~Exp*}$ is written prefix, but in the following we use infix-operator short-hand for convenience; for example, @(>) 0 1@ can be written as @0 > 1@.

@@

% ---------------------------------------------------------
\subsection{Type system}
The typing rules for Icicle are given in Figure~\ref{fig:source:type:exp}.
The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ reads ``under context $\Gamma$, the expression $\mi{Exp}$ has type $M$.''. 
The judgment form $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
The judgment form $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
The judgment form $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.
